\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-29 chapters/conclusion.tex
Layout conclusion

\chapter{Conclusion \& recommendations}
\epigraph{``The mystery of life isn't a problem to solve, but a reality to
experience.''}{ Reverend Mother Gaius Helen Mohiam in \\ \emph{Dune -- Frank
Herbert}\\ adapted from S{\o}ren Kierkegaard}
\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}

@@

\subsection{Gathered datasets}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
The dataset was presented in~\textcite{VanDenKerchove2024}.
The experiment in this dataset implemented a visual oddball \ac{bci} with six
circular targets in a hexagonal layout.
The gaze fixation of participants was carefully controlled to dissociate the
visuospatial attention of the participant and their eye gaze.
This allowed us to study the effects of gaze-independence on \ac{erp} decoding.
The CVSA-ERP dataset gives us insight into the ERP dynamics in overt, covert, and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA.
It also confirmed that the effects of covert \ac{vsa} on \ac{erp} component
amplitude hold for split \ac{vsa}.

Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
\ac{als},\ac{fa}, and stroke.
These participants exhibited varying degrees of eye motor impairment, such as
involuntary eye movements, ophthalmoplegia, and gaze fixation fatigue.
The data provide invaluable real-world evidence on how gaze-independent
\acp{bci} perform in populations that experience eye motor difficulties.
This dataset allowed us to evaluate the proposed decoding strategies in clinical settings and
highlighted the practical challenges of implementing gaze-independent \acp{bci} in these user
groups.

@@

One of the main principles of \ac{ucd} is to address the \emph{entire} user
experience.
As mentioned before, this includes measuring \emph{effectiveness}, in the form
of e.g. selection accuracy.
\Ac{ucd} goes beyond this by also assessing \emph{efficiency} and user
\emph{satisfaction}
Similarly to effectiveness, efficiency can objectively be measured as \ac{itr}
in on-line operation is a better metric.
Satisfaction, on the other hand, is a more subjective quality.

Nevertheless, optimizing user satisfaction is the penultimate goal in
application design.
Satisfaction depends on factors such as performance (in the form of efficiency
and effectiveness), comfort, and user perception.
It must be assessed through subjective questionnaires following realistic, on-line
application use\cite{Kuebler2014}.
Only focusing on evaluating effectiveness through classification performance is
a pragmatic approach, because a researcher can perform a single
data collection and iterate on the results offline.
Yet, this can be too limiting when seeking to design the most usable interface
for a specific group of users.

@@

A logical point arising from this is the question of how to set up an
effective research project that aims to optimize \ac{bci} user experience, in
the context of the topics presented in this work.
The patient-research involvement framework postulates that, ideally, individuals with
\ac{sspi} should be involved at every stage of the research project, from conceptualization
to evaluation.
A good rule of thumb is that research should solve problems \emph{with} the
users, not \emph{for} them, implying that research projects should start from
the needs of the \ac{bci} user population.

@@

This approach is tied to another valuable \ac{ucd} principle: encouraging early and active
involvement of users.
Unfortunately, it also might one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and what does not work for them, and where research efforts
should be focused.

>>>>

\chapter{Conclusion \& recommendations}
\epigraph{``The mystery of life isn't a problem to solve, but a reality to
experience.''}{The Reverend Mother Gaius Helen \\ Mohiam in \emph{Dune -- Frank
Herbert}, \\ originally from S{\o}ren Kierkegaard}
\publicationnote{\Cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}

@@

\subsection{Gathered datasets}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
26.38$\pm$3.15 years.
The dataset was presented in~\textcite{VanDenKerchove2024}.
The experiment in this dataset implemented a visual oddball \ac{bci} with six
circular targets in a hexagonal layout.
The gaze fixation of participants was carefully controlled to dissociate the
visuospatial attention of the participant and their eye gaze.
This allowed us to study the effects of gaze-independence on \ac{erp} decoding.
The CVSA-ERP dataset gives us insight into the ERP dynamics in overt, covert, and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA.
It also confirmed that the effects of covert \ac{vsa} on \ac{erp} component
amplitude hold for split \ac{vsa}.

Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
\ac{als}, \ac{fa} and stroke.
These participants exhibited varying degrees of eye motor impairment, such as
involuntary eye movements, ophthalmoplegia, and gaze fixation fatigue.
The data provide invaluable real-world evidence on how gaze-independent
\acp{bci} perform in populations that experience eye motor difficulties.
This dataset allowed us to evaluate the proposed decoding strategies in clinical settings and
highlighted the practical challenges of implementing gaze-independent \acp{bci} in these user
groups.

@@

One of the main principles of \ac{ucd} is to address the \emph{entire} user
experience.
As mentioned before, this includes measuring \emph{effectiveness}, in the form
of e.g. selection accuracy.
\Ac{ucd} goes beyond this by also assessing \emph{efficiency} and user
\emph{satisfaction}
Similarly to effectiveness, efficiency can objectively be measured as the \ac{itr}
achieved in -- importantly -- on-line operation.

Satisfaction, on the other hand, is a more subjective quality.
Nevertheless, optimizing user satisfaction is the penultimate goal in
application design.
Satisfaction depends on factors such as performance (in the form of efficiency
and effectiveness), comfort, and user perception.
It must be assessed through subjective questionnaires following realistic, on-line
application use\cite{Kuebler2014}.
Only focusing on evaluating effectiveness through classification performance is
a pragmatic approach, because a researcher can perform a single
data collection and iterate on the results offline.
Yet, this can be too limiting when seeking to design the most usable interface
for a specific group of users.

@@

A logical point arising from this is the question of how to set up an
effective research project that aims to optimize \ac{bci} user experience, in
the context of the topics presented in this work.
Another valuable principle of \ac{ucd} principle points the way: encouraging early and active
involvement of end-users.
This entails that, ideally, individuals with \ac{sspi} should be involved at
every stage of the research project, from conceptualization
to evaluation.
A good rule of thumb is that research should solve problems \emph{with} the
users, not \emph{for} them, implying that research projects should start from
the needs of the \ac{bci} user population.

@@

Unfortunately, this approach might also be one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
It can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and what does not work for them, and where research efforts
should be focused.

### 2024-11-29 chapters/patients.tex
Layout conclusion

\chapter{Case studies}
\label{sec:patients}
\epigraph{I knew there would be a lot of unexpected issues involved with
patient experiments, but I would never have guessed that there would be baby
goats in the experiment room\ldots}{Juliette}

@@

\section{Methods}
\subsection{Recruitment}
Participants were recruited across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item be unable to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
criterion~\ref{item:patients/exclusion/vision}, one person recovering from
\ac{tbi} on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
stroke based
on~\ref{item:patients/exclusion/medical}.
One further person recovering from a stroke was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed with bulbar-onset \ac{als}, three with
\ac{fa} an three were recovering from stroke.
We refer to \cref{sec:gaze-independent/gaze-dependence/incidence} for a short
description of these conditions.
Table~\ref{tab:patients/patients} lists the included participants and their
diagnoses.

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
%\todo{include years since diagnosis}

\subsection{Visual skills and eye tracking and eye motor examination}

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  \caption[Distribution of the recorded gaze position.]{%
    Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
    conditions.
    Crosshairs represent stimulus positions, with the orange ones cued during
    the given condition.
    Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
    near the middle of the screen, and PC4 near the bottom.
  }
  \label{fig:patients/gaze}
}

>>>>

\chapter{Case studies}
\label{sec:patients}
\epigraph{``I knew there would be a lot of unexpected issues involved with
patient experiments, but I never would have guessed that there would be a baby
goat in the experiment room\ldots''}{Juliette}

@@

\section{Methods}
\subsection{Recruitment}
Participants were recruited across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item had a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item had a predisposition to or a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item had a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item were unable to understand the experiment instructions and cooperate,
  \item had any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 individuals were contacted. Of these, which one person with
\ac{ms} was excluded based on criterion~\ref{item:patients/exclusion/vision}.
One person recovering from \ac{tbi} was excluded based on both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
stroke based on~\ref{item:patients/exclusion/medical}.
One further person recovering from a stroke was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed with bulbar-onset \ac{als}, three with
\ac{fa} an three were recovering from stroke.
We refer to \cref{sec:gaze-independent/gaze-dependence/incidence} for a short
description of these conditions.
Table~\ref{tab:patients/patients} lists the included participants and their
diagnoses.
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, W: classification according
  to~\textcite{Wolpaw2006}\footnotemark[1], KB: classification according
  to~\textcite{Kuebler2008}\footnotemark[2].
  }
  \label{tab:patients/patients}
\end{table}
\subsection{Visual skills and eye tracking and eye motor examination}
\footnotetext[1]{See \cpageref{fn:bci/wolpawcrit} \cref{fn:bci/wolpawcrit}.}
\footnotetext[2]{``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech. \elide Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}}
%\todo{include years since diagnosis}

@@

\begin{table}[t]
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

@@

\fullpagefig[anchor.north]{%
  %\includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
  %\includegraphics[width=\textwidth]{figures/patients/fig_gaze.pdf}
  \inputpgf{figures/patients}{fig_gaze.pgf}
}{%
  \caption[Distribution of the recorded gaze position.]{%
    Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
    conditions.
    Crosshairs represent stimulus positions, with the orange ones cued during
    the given condition.
    Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
    near the middle of the screen, and PC4 near the bottom.
  }
  \label{fig:patients/gaze}
}

### 2024-11-29 chapters/patients_table.tex
Layout conclusion

\begin{tabular}{llllllllr}
  \toprule
  \textbf{ID}  & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
  \textbf{Diagnosis}
  & \textbf{Speech}     & \textbf{Trach.} & \textbf{Communication}          &
  \textbf{Cls.} \\ \midrule
  PA1 & 58  & M   & L     & bulbar-onset \acs{als} & anarthric  & no          & tablet                 & 3  \\
  PB1 & 41  & M   & L     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PB2 & 43  & F   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PB4 & 48  & M   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PC2 & 43  & M   & R     & brainstem stroke & anarthric  & yes         & \makecell[l]{prompting\\+eye movement} & 2 \\
  PC3 & 43  & F   & R     & brainstem stroke & anarthric  & yes         & letterboard            & 2 \\
  PC4 & 54  & M   & R     & \makecell[l]{left cerebellar stroke \\ (trombosis of the basilar artery)} & anarthric  & yes & letterboard & 2 \\
  \bottomrule
\end{tabular}

>>>>

\sffamily
\small
\begin{tabular}{@{}llrrrlllrr@{}}
  \toprule
  \textbf{ID}  & \textbf{Diagnosis} & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
  \textbf{Speech} & \textbf{Trach.} & \textbf{Communication} &
  \textbf{W} & \textbf{KB} \\ \midrule
  PA1 & bulbar-onset \acs{als} & 58  & M   & L     & anarthric  & no          & tablet                 & 3  & 4 \\
  PB1 & \acs{fa} & 41  & M   & L     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB2 & \acs{fa} & 43  & F   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB4 & \acs{fa} & 48  & M   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PC2 & brainstem stroke & 43  & M   & R     & anarthric  & yes         &  \makecell[l]{prompting\\+eye movement} & 2 & 4  \\
  PC3 & brainstem stroke & 43  & F   & R     & anarthric  & yes         & letterboard            & 2  & 3 \\
  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  basilar \\ artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
  \bottomrule
\end{tabular}

### 2024-11-29 chapters/stbf_struct.tex
Layout conclusion

\chapter{Kronecker-structured linear decoding}
\label{sec:stbf-struct}
\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
\epigraph{%
  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
  With these hundred coins, I paid for the beast and received three coins in change.
  Of the three, I returned one coin each to my companions, keeping one for myself.
  Now, I owe each friend forty-and-nine coins, a total debt of
  ninety-and-eight. And with mine one remaining, the
  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
  Tell me then, kind stranger, whither hath the missing coin gone?''
}{%
  Ancient Algerian riddle
}
% Kronecker-structured LDA ====================================================

>>>>

\chapter{Kronecker-structured linear decoding}
\label{sec:stbf-struct}
\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
\epigraph{%
  I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
  With these hundred coins, I paid for the beast and received three coins in change.
  Of the three, I returned one coin each to my companions, keeping one for myself.
  Now, I owe each friend forty-and-nine coins, a total debt of
  ninety-and-eight. And with mine one remaining, the
  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
  Tell me then, kind stranger, whither hath the missing coin gone?
}{%
  Ancient Algerian riddle
}
% Kronecker-structured LDA ====================================================
